A Bayesian statistical model is able to predict target-by-target selection behaviour in a human foraging task

### 2022-10-14 Analysis/index_revisions_2.Rmd
work on more revisions

---
title: 'SVG: The Paper'
author: "Alasdair Clarke, Amelia Hunt &  Anna Hughes"
date: "13/04/2022"
output:
  tufte::tufte_html:
    number_sections: true
    fig_height: 4
    bibliography: literature.bib
    link-citations: yes  
---

@@

# for saving graphs for paper
theme_set(ggthemes::theme_tufte() + 
            theme(plot.background = element_rect(fill = "#ffffff", colour = "#ffffff")))

@@

```{r, fig.cap = "(*left*) The average weight assigned to each selected target by our model. (*right*) The proportion of trials the item with the largest assigned weight was selected by the participant.  Each dot shows data from an individual participant in a condition and the shaded region indicates the interval in which we expect 67% of participants to fall."}

@@

ggsave("../Figures/qjep_preds.pdf", 
       plt_b2 / (plt1 + plt2) + 
       plot_layout(guides = "collect") & theme(legend.position = 'bottom'),
       width = 10, height = 6)

@@

### Individual Differences

@@

plt_b <- ggplot(filter(fit_plt, parameter == "bP"), 
                aes(value, accuracy, colour = condition)) + 
  geom_point(aes(shape = is_outlier)) +
  geom_smooth(data = filter(fit_plt, is_outlier == 0, parameter == "bP"), method = "lm", formula = y ~ x, se = FALSE) + 
  scale_shape_manual(values = c(19, 4), guide = "none") + 
  theme(legend.position = "none") + 
   scale_y_continuous("model accuracy",  
                     limits = c(0.35, 0.8), breaks = seq(0.3, 0.8, 0.1))

@@

### Shorter trials are more predictable

>>>>

---
title: 'SVG: The Paper'
author: "Alasdair Clarke, Amelia Hunt &  Anna Hughes"
date: "13/04/2022"
output:
  tufte::tufte_handout:
    number_sections: true
    fig_height: 4
---

@@

# for saving graphs for paper
theme_set(ggthemes::theme_tufte(base_size = 16) + 
            theme(plot.background = element_rect(fill = "#ffffff", colour = "#ffffff")))

@@

```{r, fig.cap = "Left: The average weight assigned to each selected target by our model. Right: The proportion of trials the item with the largest assigned weight was selected by the participant.  Each dot shows data from an individual participant in a condition and the shaded region indicates the interval in which we expect 67% of participants to fall."}

@@

ggsave("../Figures/qjep_preds.pdf", 
       plt_b2 / (plt1 + plt2) + 
       plot_layout(guides = "collect") & theme(legend.position = 'bottom'),
       width = 10, height = 10)

@@

## Individual Differences

@@

plt_b <- ggplot(filter(fit_plt, parameter == "bP"), 
                aes(value, accuracy, colour = condition)) + 
  geom_point(aes(shape = is_outlier)) +
  geom_smooth(data = filter(fit_plt, is_outlier == 0, parameter == "bP"), method = "lm", formula = y ~ x, se = FALSE) + 
  scale_shape_manual(values = c(19, 4), guide = "none") + 
  xlab("bP parameter value") +
  theme(legend.position = "none") + 
   scale_y_continuous("model accuracy",  
                     limits = c(0.35, 0.8), breaks = seq(0.3, 0.8, 0.1))

@@

## Shorter trials are more predictable

### 2022-10-14 manuscriptRevisions2.tex
work on more revisions

We use a previously published dataset to explore how well our model can account for trial-level behaviour: the visual foraging data from \cite{clarke2022}. This is an attractive dataset for our needs as it is very close to the ``classic'' visual foraging paradigm \citep{kristjansson2014}, yet with a substantially larger sample of participants. (The key difference between paradigms is that \cite{kristjansson2014} used an ipad and finger foraging, while \cite{clarke2022} used a desktop computer and mouse clicks.) Furthermore, we have already demonstrated that our model can capture individual differences in foraging behaviour \citep{clarke2022}. As such, we will only give a brief overview of the data here and refer the reader to these earlier papers for more details.

@@

We will make use of the foraging model from \cite{clarke2022}. This treats foraging as a sampling without replacement process in which each item $i$ has probability $p_i$ of being selected as the next target. The $p_i$ depend on four parameters:

@@

To assess how well our model can account for behaviour at the level of individual trials, we start by stepping through each trial in the data from \cite{clarke2022}. For each target selection\footnote{Note: for the reasons discussed above, we ignore the initial target selection for now, as the model has very little to say about it. We return to this issue in section 4.}, in each trial, we can estimate how likely each of the remaining items are to be selected using the parameters from our posterior model. As can be seen in Figure \ref{fig:cal}, the model is putting somewhere between 25\% and 75\% of the weight on the target that is selected next, easily outperforming a chance $1/n$ baseline. We can also see  that the model is well calibrated in that the probabilities assigned to the most likely target manage to capture how often that target is actually selected. However, we can see a decrease in accuracy when it comes to how often ``runner-up`` candidate items are selected as our model appears to systematically undervalue these (presumably by putting too much of the probabilistic weight on low chance items).

@@

One weakness of the foraging model by \cite{clarke2022foraging} is that it makes little attempt to predict which item will be selected first. In this section we aim to improve this by modelling participant bias (or preference) in their choice of initial target selection. Our approach is inspired by previous work on the central bias in scene viewing \cite{clarke_tatler2014, clarke2017}. These studies fit truncated Gaussian distributions to the  $(x,y)$ coordinates of fixation locations. Here we will do similar, and fit beta distributions\footnote{We use beta distributions here as ADFC couldn't work out the syntax for specifying a mixture model using truncated Gaussians in Stan. But, in hindsight, beta distributions are probably a better choice anyway!} to locations of the initial target selections in the visual foraging task.

@@

This reduces the number of parameters to eight (to specify the two Beta distributions) and then one $\lambda$ value per participant (= 66 in our dataset). While the formulae in Equations \ref{eq:beta2x} and \ref{eq:beta2y} appear more complex than those in Equations \ref{eq:beta1x} and \ref{eq:beta1y}, the model requires less than a quarter as many free parameters!

@@

A benefit of our approach is that it is relatively easy to interrogate the model to try to understand why prediction accuracy is higher for some participants and trials than others. The model is generally better at feature search compared to conjunction search, and also seems to be better at predicting participants with a stronger proximity bias. The model also seems to assign more weight to 'runner-up' items than real humans do, perhaps suggesting a 'peakier' choice distribution, but with fatter tails, for real foragers, with only a few nearby targets seriously considered, but with some low probability of selecting a further away target.

>>>>

We use a previously published dataset to explore how well our model can account for trial-level behaviour: the visual foraging data from Clarke and colleagues \cite{clarke2022}. This is an attractive dataset for our needs as it is very close to the ``classic'' visual foraging paradigm conducted by Kristjánsson and colleagues \citep{kristjansson2014}, yet with a substantially larger sample of participants. (The key difference between paradigms is that \cite{kristjansson2014} used an ipad and finger foraging, while \cite{clarke2022} used a desktop computer and mouse clicks.) Furthermore, we have already demonstrated that our model can capture individual differences in foraging behaviour \citep{clarke2022}. As such, we will only give a brief overview of the data here and refer the reader to these earlier papers for more details.

@@

We will make use of the foraging model from Clarke et al (2022) \cite{clarke2022foraging}. This treats foraging as a sampling without replacement process in which each item $i$ has probability $p_i$ of being selected as the next target. The $p_i$ depend on four parameters:

@@

To assess how well our model can account for behaviour at the level of individual trials, we start by stepping through each trial in the data from Clarke and colleagues \cite{clarke2022}. For each target selection\footnote{Note: for the reasons discussed above, we ignore the initial target selection for now, as the model has very little to say about it. We return to this issue in section 4.}, in each trial, we can estimate how likely each of the remaining items are to be selected using the parameters from our posterior model. As can be seen in Figure \ref{fig:cal}, the model is putting somewhere between 25\% and 75\% of the weight on the target that is selected next, easily outperforming a chance $1/n$ baseline. We can also see  that the model is well calibrated in that the probabilities assigned to the most likely target manage to capture how often that target is actually selected. However, we can see a decrease in accuracy when it comes to how often ``runner-up`` candidate items are selected as our model appears to systematically undervalue these (presumably by putting too much of the probabilistic weight on low chance items).

@@

One weakness of the foraging model by Clarke et al (2022) \cite{clarke2022foraging} is that it makes little attempt to predict which item will be selected first. In this section we aim to improve this by modelling participant bias (or preference) in their choice of initial target selection. Our approach is inspired by previous work on the central bias in scene viewing \cite{clarke_tatler2014, clarke2017}. These studies fit truncated Gaussian distributions to the  $(x,y)$ coordinates of fixation locations. Here we will do similar, and fit beta distributions\footnote{We use beta distributions here as it is difficult to specify a mixture model using truncated Gaussians in Stan. However, beta distributions may well be a more appropriate choice.} to locations of the initial target selections in the visual foraging task.

@@

This reduces the number of parameters to eight (to specify the two Beta distributions) and then one $\lambda$ value per participant (= 66 in our dataset). While the formulae in Equations \ref{eq:beta2x} and \ref{eq:beta2y} appear more complex than those in Equations \ref{eq:beta1x} and \ref{eq:beta1y}, the model requires less than a quarter as many free parameters.

@@

A benefit of our approach is that it is relatively easy to interrogate the model to try to understand why prediction accuracy is higher for some participants and trials than others. The model is generally better at feature search compared to conjunction search, and also seems to be better at predicting participants with a stronger proximity bias. The model also seems to assign more weight to 'runner-up' items than real humans do. Real foragers thus seem to seriously consider only a few nearby targets (although they do still have some low probability of selecting a further away target), in contrast to the model, in which the item weights assigned fall off more gradually with distance.
